Data-driven Reduction of Multiscale Stochastic Dynamical Systems \thanks{This work was supported by ....}

### 2014-08-31 reducible_SDE/paper/SIADS_draft.tex
added more it taylor calc

Add a general formulation of a reducible system (with a general drift, and decoupled diffusions) and Dror/Kasmiskii's conditions.
(We assume that the saturation time of fast scale  of the SDE is smaller than the other relevant timescales.)
\begin{equation} 
\begin{aligned}
dx &= a(x,y) dt + dW_1\\
dy &= -\frac{b(x,y)}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
\end{aligned}
\end{equation}

Given a system of SDEs with separation of timescales, we want to recover slow variable(s) even when the fast variables are have large variations.

We consider systems that can be written as deterministic nonlinear functions $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^m$ of the following system of SDEs (in the general form, in particular, this one):
\begin{equation} \label{eq:fast_slow_SDE}
\begin{aligned}
dx &= adt + dW_1\\
dy &= -\frac{y}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
\end{aligned}
\end{equation}
where the diffusion terms of the fast and slow variables are decoupled.

So $x$ is the slow variable, and $y$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
%
We would like to recover $x$ using data-driven techniques.

We would like to note that the ratio of the drift and diffusion terms in the $y$ equation is essential.
%
We need the square of the diffusivity to be of the same order as the drift.
%
If the diffusivity is too large, the equilibrium measure of $y$ will be unbounded.
%
Conversely, if the diffusivity is too small, the equilibrium measure of $y$ will go to 0.

We will consider two specific examples. 
\begin{enumerate}
\item fat strip
\item half moon
\end{enumerate}
%
We will show in both cases how we can recover the slow variables using data-driven techniques.

@@

Assume we want to locally recover the slow variables. 
%
To do so, we need to build a metric that is invariant to the fast variables. 
%
This is typically done by averaging out the fast variables. 
%
We propose to do this using the Mahalanobis distance, which is a local PCA-type approach that inverts the local covariance matrix of the data, thereby creating a space with normalized diffusion terms.
%
It can be shown that the Mahalanobis distance locally approximates: 
\begin{equation}
\| f(x_1, y_1) - f(x_2, y_2) \|^2_M = \| X_1 - X_2 \|^2_2 + \mathcal{O}(\| f(x_1, y_1) - f(x_2, y_2) \|^4_2)
\end{equation}
where $X$ comprises variables which have uncoupled noise with unit variance,
i.e., the variables $X^1, \dots, X^n$ are governed by the following SDEs:
\begin{equation} \label{eq:NIV_formulation}
dX^i = a(X) dt + dW^i
\end{equation}
where $X = \begin{bmatrix} X^1 & X^2 & \cdots & X^n \end{bmatrix}^T$.
%
In particular for the case TODO, consider the change of variable $z = y\sqrt{\epsilon}$. This change of variables gives the following system:
\begin{equation}
\begin{aligned}
dx &= adt + dW_1\\
dz &= -\frac{z}{\epsilon} dt +  dW_2
\end{aligned}
\end{equation}
Therefore, $z$ is a stochastic variable whose diffusion is unity, and $X = \begin{bmatrix} x \\ z \end{bmatrix}$.

Note that for this specific case
\begin{equation}
\| X_1 - X_2 \|^2_2 = \|x_1 - x_2 \|^2_2 + \epsilon \|y_1 - y_2 \|^2_2.
\end{equation}
Remarkably, it implies two features.
%
One, it rescales that data (in the sense that the variables have unit diffusion terms) so that the fast variables are collapsed and become (``epsilon") small as observed in the second term in the right hand side. 
%
Two, it inverts the function $f$ (locally) and allows for the approximation of the Euclidean distance between samples of the slow variables in the ``decoupled" SDE space.

%These same conditions for uncoupled noises with unit variance will allow us to recover the slow variable(s) from data with multiple timescales. 
%
%

%To get a parameterization of NIV, we require $\| X_2 - X_1\|_2$.
%
%However, we only have access to $Y = f(X)$.
%
%Assuming that we have a local approximation to $\| X_2 - X_1\|$ from the measurements $Y$, we can then use DMAPS to uncover a parameterization of $X$ from the estimates of the pairwise distances.

\section{DMAPS/Graph-Based/Kernel-based method}

Description of DMAPS........

@@

See schematic figure.

Our approximation of the pairwise distances will rely on the Taylor expansion of the measurement function $f$. 
%
We will then {\em empirically} estimate the first-order term in this expansion using simulation bursts to estimate the local covariance.  
%
Accordingly, we will address the accuracy of our method from two standpoints: (1) the accuracy of the Taylor expansion, and (2) the accuracy of the covariance estimation (see Figure...).
%
We will present both analytical results for the error bounds, as well as an empirical methodology to set the appropriate parameters for our method to accurately recover the intrinsic slow variable(s).

@@

>>>>

We consider the following two-scale SDE. 
\begin{equation} \label{eq:general_SDE}
\begin{aligned}
dx_i &= a_i(x_1, \dots, x_n) dt + dW_i, \: 1 \le i \le m \\
dx_i &= -\frac{a_i(x_1, \dots, x_n)}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_i , \: m+1 \le i \le n
\end{aligned}
\end{equation}
where $W_i$ are independent Brownian motions, and $\epsilon \ll 1$.
%
Therefore, \eqref{eq:general_SDE} defines an $n$-dimensional stochastic system with $m$ slow variables and $n-m$ fast variables. 
%
We would like to note that the ratio of the drift and diffusion terms is essential.
%
We need the square of the diffusivity to be of the same order as the drift.
%
If the diffusivity is too large, the equilibrium measure will be unbounded.
%
Conversely, if the diffusivity is too small, the equilibrium measure will go to 0.

Assuming that the functions $a_i$ are bounded and Lipschitz, and we can find
\begin{equation}
\mathbb{E} \left| \frac{1}{T} \int_t^{t + T} a_i(x_1, \dots, x_m, X_{m+1}(s), \dots, X_n(s)) ds - \overline{a} (x_1, \dots, x_m) \right| < \kappa_i (T) 
\end{equation}
for $1 \le i \le m$, where $\kappa_i((T) \rightarrow 0 $ as $T \rightarrow \infty$, then the averaging principle \cite{...} states that we can write an effective equation in the slow variables $x_1, \dots, x_m$. 
%
We will consider systems that can be written as deterministic (perhaps nonlinear) bi-Lipshitz functions $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^d$ of the SDE in \eqref{eq:general_SDE}.
%
Our goal is to recover the slow variables $x_1, \dots, x_m$ using data-driven techniques.

For illustrative purposes, we will consider the following two-dimensional SDE as a specific example of \eqref{eq:general_SDE}. 
\begin{equation} \label{eq:fast_slow_SDE}
\begin{aligned}
dx_1 &= adt + dW_1\\
dx_2 &= -\frac{x_2}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
\end{aligned}
\end{equation}
%
where $a$ is a constant of order 1. 
%
So $x_1$ is the slow variable, and $x_2$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
%
We would like to recover $x_1$ using data-driven techniques.

We will consider two specific examples. 
%
In the first example, our data will form a strip where the fast and slow remain uncoupled,
\begin{equation}
\mathbf{f_1}(x_1, x_2) = \begin{bmatrix} x_1 \\ x_2 \end{bmatrix}
\end{equation}

In the second example, our data will be warped into half-moon shapes
\begin{equation}
\mathbf{f_2}(x_1, x_2) = 
\begin{bmatrix} 
x_1 + x_2^2 \\
x_2
\end{bmatrix}
\end{equation}
%
We will show in both cases how we can recover the slow variables using data-driven techniques.

@@

To locally recover the slow variables, we need to build a metric that is invariant to the fast variables. 
%
Typically, this is done by averaging out the fast variables. 
%
We propose to do this using the Mahalanobis distance, which is a local PCA-type approach that inverts the local covariance matrix of the data, thereby creating a space with normalized diffusion terms.
%
It can be shown that the Mahalanobis distance locally approximates: 
\begin{equation} \label{eq:mahalanobis}
\| \mathbf{f}(Y_1) - \mathbf{f}(Y_2) \|^2_M = \| Y_1 - Y_2 \|^2_2 + \mathcal{O}(\| \mathbf{f}(Y_1) - \mathbf{f}(Y_2) \|^4_2)
\end{equation}
where the components of $Y$ have uncoupled noise with unit variance,
i.e., the variables $y_1, \dots, y_n$ are governed by the following SDEs:
\begin{equation} \label{eq:NIV_formulation}
dy_i = b(y) dt + dW_i
\end{equation}
and $Y = \begin{bmatrix} y_1 & y_2& \cdots & y_n \end{bmatrix}^T$.

In particular, for SDEs of the form in \eqref{eq:general_SDE}, consider the transformation
\begin{equation} \label{eq:general_rescale}
\begin{aligned}
x_i = y_i, \: &1 \le i \le m \\
x_i = \frac{y_i}{\sqrt{\epsilon}} \: &1 \le i \le m
\end{aligned}
\end{equation}

Then, \eqref{eq:general_SDE} becomes
\begin{equation} 
\begin{aligned}
dy_i &= b_i(y_1, \dots, y_n) dt + dW_i, \: 1 \le i \le m \\
dy_i &= -\frac{b_i(y_1, \dots, y_n)}{\sqrt{\epsilon}} dt + dW_i , \: m+1 \le i \le n
\end{aligned}
\end{equation}
where $b_i(y_1, \dots, y_n) = a_i (y_1, \dots, y_m, \frac{y_{m+1}}{\sqrt{\epsilon}}, \dots, \frac{y_n}{\sqrt{\epsilon}})$.
%
From \eqref{eq:general_rescale}, we can see that if $x_i = \mathcal{O}(1)$, then $y_i = \mathcal{O}({\sqrt{\epsilon}})$ for $ m+1 \le i \le n$, and 
\begin{equation}
\| Y_2 - Y_1 \|^2_2 = \sum_{i=1}^m \| y_{i,2} - y_{i,1} \|^2 = \sum_{i=1}^m \|x_{i,2} - x_{i,1} \|^2 + \epsilon \sum_{i=m+1}^n \|x_{i,2} - x_{i,1}\|^2
\end{equation}
%
Remarkably, this metric induces two important features. 
%
One, rescaling the data so that all variables have unit diffusion terms means that fast variables are collapsed and become ``epsilon" small, as observed in the second term in the right hand side. 
%
Two, the righthandside of \eqref{eq:mahalanobis} is invariant (to fourth order) to the function $\mathbf{f}$ and allows for the approximation of the Euclidean distance between samples of the slow variables in the ``decoupled" SDE space.

For the specific example \eqref{eq:fast_slow_SDE}, the change of variables gives the following system:
\begin{equation}
\begin{aligned}
dy_1 &= adt + dW_1\\
dy_2 &= -\frac{y}{\epsilon} dt +  dW_2
\end{aligned}
\end{equation}
Therefore, $y_2$ is a stochastic variable whose diffusion is unity, and 
\begin{equation}
\| X_1 - X_2 \|^2_2 = \|x_1 - x_2 \|^2_2 + \epsilon \|y_1 - y_2 \|^2_2.
\end{equation}

\section{Diffusion Maps for Global Parameterization}

Our goal is to extract a {\em global} parameterization of the data that respects the slow variables. 
%
We will use diffusion maps, a kernel-based manifold learning technique, to extract a global parameterization using the local distances that we described in the previous section. 
%
Given data $z_1, \dots, z_N \in \mathbb{R}^d$, we first construct the matrix $W \in \mathbb{R}^{N \times N}$, where 
\begin{equation}
W_{ij} = \exp \left( -\frac{\|z_i - z_j \|^2}{\sigma_{kernel}^2} \right)
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm, and $\sigma_{kernel}$ is the kernel scale and denotes a characteristic distance within the data set (we often set $\sigma_{kernel}$ to be the median of the pairwise distances). 
%
In our applications, we will use the Mahalanobis distance \eqref{eq:mahalanobis} as the norm between points, so that we can recover the slow variables.

We then construct the diagonal matrix $D \in \mathbb{R}^{N \times N}$, with 
\begin{equation}
D_{ii} = \sum_{j=1}^N W_{ij}
\end{equation}
%
We compute the eigenvalues $\lambda_0, \dots, \lambda_{N-1}$ and eigenvalues $\phi_0, \dots, \phi_{N-1}$ of the matrix $A = D^{-1}W$, and order them such that $|\lambda_0| \ge |\lambda_1| \ge \dots \ge |\lambda_{N-1}|$. 
%
$\phi_0$ is a constant trivial eigenvector; the next few eigenvectors give a parameterization/embedding coordinates for the data.

@@

Our approximation of the pairwise distances will rely on the Taylor expansion of the measurement function $f$. 
%
We will then {\em empirically} estimate the first-order term in this expansion using simulation bursts to estimate the local covariance.  
%
Accordingly, we will address the accuracy of our method from two standpoints: (1) the accuracy of the Taylor expansion, and (2) the accuracy of the covariance estimation (see Figure~\ref{fig:schematic}).
%
We will present both analytical results for the error bounds, as well as an empirical methodology to set the appropriate parameters for our method to accurately recover the intrinsic slow variable(s).

@@

\begin{equation}
\begin{aligned}
f( X_\tau) = & f( X_\rho) \\
&+ \sum_{j=1}^d \left. \frac{\partial f}{\partial x^j} \right|_{X_\rho}  \int_{\rho}^{\tau}  dW_s^j \\ &
+  \left. \left( \sum_{k=1}^d a^k \frac{\partial f}{\partial x^k} + \frac{1}{2} \sum_{j=1}^d \frac{\partial^2 f}{\partial {x^j}^2} \right) \right|_{X_\rho} \int_{\rho}^{\tau} ds + \sum_{j=1}^m \sum_{k=1}^m \int_{\rho}^{\tau} \int_{\rho}^{s_2} \frac{\partial^2 f}{\partial x^j \partial x^k} dW_{s_1}^j dW_{s_2}^k \\
& + \sum_{j=1}^d  \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{k=1}^d a^k \frac{\partial^2 f}{\partial x^j \partial x^k} + \frac{1}{2} \sum_{k=1}^d \frac{\partial^3 f}{\partial x^j \partial {x^k}^2} \right) dW_{s_1}^j ds_2 \\
& + \sum_{j=1}^d  \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{k=1}^d a^k \frac{\partial^ f}{\partial x^k \partial x^j} + \frac{1}{2} \sum_{k=1}^d  \frac{\partial^3 f}{\partial {x^k}^2 \partial x^j} \right) ds_1 dW_{s_2}^j  \\
& + \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{j=1}^d \sum_{k=1}^d \left[ a^j \frac{\partial}{\partial x^j}\left( a^k \frac{\partial f}{\partial x^k} \right) + \frac{1}{2} \frac{\partial^2}{\partial {x^j}^2} \left( a^k \frac{\partial f}{\partial x^k} \right) + \frac{1}{2} a^j \frac{\partial^3 f}{\partial x^j \partial {x^k}^2} + \frac{1}{4} \frac{\partial^4 f}{\partial {x^j}^2 \partial {x^k}^2} \right] \right)  ds_1 ds_2
\end{aligned}
\end{equation}
